---
title: "Week 2 · Class 2: Practice Exercises"
subtitle: "Working with Real Data & Summary Statistics"
date: "2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(formatR)
# When running chunks interactively, set the working directory to this file's
# folder so the relative data/ paths resolve. (During render, knitr already
# uses the document's directory, so this is skipped.)
if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  .doc_path <- rstudioapi::getSourceEditorContext()$path
  if (length(.doc_path) == 1 && nzchar(.doc_path)) setwd(dirname(.doc_path))
}
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Segment A — Working with Real Data
```{r echo=F}
library(knitr)
library(formatR)
library(tidyverse)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Non-AI Exercises

## 1. Working with Missing Data

### 1.1 The Challenge of Missing Data

Why can't we use `== NA` to check for missing values in R? What does this tell us about how R thinks about missing data?

Answer:

### 1.2 Code Detective: Missing Data

Explain what's wrong with this code and how to fix it:

```r
data %>%
  filter(age == NA)
```

Problem: _____
Solution: _____



## 2. The Pipe Operator

### 2.1 Pipe vs No Pipe

These two pieces of code do the same thing. Which is easier to read and why?

```r
# Version A
arrange(select(filter(congress, party_code == "Republican"), bioname, nominate_dim1), nominate_dim1)

# Version B  
congress %>%
  filter(party_code == "Republican") %>%
  select(bioname, nominate_dim1) %>%
  arrange(nominate_dim1)
```

Answer: _____

## 3. Understanding DW-NOMINATE Scores

### 3.1 The Scale

DW-NOMINATE scores range from -1 to +1. What do these values represent in terms of political ideology? If a member of Congress has a score of -0.8, what does this tell you about their political positions?

Answer:

### 3.2 Interpreting Changes

If the average DW-NOMINATE score for Democrats has moved from -0.3 in 1980 to -0.4 in 2020, what does this suggest about the party? What about if Republicans moved from +0.3 to +0.5?

Answer:

## 4. Advanced Tidyverse Operations

### 4.1 The %in% Operator

Explain what this code does and when you would use the `%in%` operator instead of `==`:

```r
congress %>%
  filter(state_abbrev %in% c("CA", "TX", "NY", "FL"))
```

Answer:

### 4.2 Select Helpers

Match each `select()` helper function with what it does:

a) `starts_with("vote")`     1. Selects columns containing "district" anywhere
b) `ends_with("_id")`        2. Selects all remaining columns
c) `contains("district")`     3. Selects columns starting with "vote"
d) `everything()`            4. Selects columns ending with "_id"

Matches: a = _____, b = _____, c = _____, d = _____

## 5. Building Complex Analyses

### 5.1 Step-by-Step Approach

Why is it important to build your analysis step by step rather than writing complex chains all at once? Give two specific benefits.

Answer:

## 6. Data Exploration Best Practices

### 6.1 Understanding Your Data

Why is it important to use functions like `count()`, `summary()`, and `glimpse()` before doing complex analysis? What might happen if you skip this step?

Answer:

## 7. The NOT Operator

### 7.1 Using the NOT Operator

What will each of these expressions return if `x = 5` and `y = NA`:

a) `!TRUE` = _____
b) `!(x > 3)` = _____
c) `!is.na(x)` = _____
d) `!is.na(y)` = _____

### 7.2 Practical Application

Write the filter condition to keep only rows where the `vote_share` column is NOT missing:

Answer: `filter(_____)`

## 8. Multiple Sorting

### 8.1 Understanding arrange() with Multiple Variables

What does this code do, and in what order will the results appear?

```r
congress %>%
  arrange(state_abbrev, desc(nominate_dim1))
```

Answer:

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, calculate the mean nominate_dim1 by party_code. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

## 9. Congressional Ideology Analysis

**Dataset: data/HSall_members.csv**

**Description**: DW-NOMINATE scores for all members of Congress throughout history.

**Variables**:
- `congress`: Congress number (int)
- `chamber`: House or Senate (chr)
- `state_abbrev`: State abbreviation (chr)
- `party_code`: Political party (chr)
- `bioname`: Member name (chr)
- `nominate_dim1`: Liberal-conservative score, -1 to 1 (dbl)
- `nominate_dim2`: Second dimension score (dbl)
- `nominate_geo_mean_probability`: Prediction accuracy (dbl)

### 9.1 Loading and Initial Exploration

```{r}
# Load the dataset
congress <- read_csv("data/HSall_members.csv")

# Explore the DW-NOMINATE data
glimpse(congress)
```

This dataset contains ideological scores for all members of Congress. Work with Claude to understand what the nominate_dim1 variable represents.

### 9.2 Party Polarization Over Time

Ask Claude to help you analyze how the ideological distance between Democrats and Republicans in the most recent Congress.

```{r}

```

### 9.3 State-Level Analysis

Work with Claude to find which state has the most ideologically diverse congressional delegation in the current Congress.

```{r}

```

## 10. School Climate and Outcomes

**Dataset: data/school_climate.csv**

**Description**: Student-level data nested in 500 schools measuring climate, discipline, academics.

**Variables**:
- `student_id`: Unique student identifier (int)
- `school_id`: School identifier (int)
- `grade_level`: 9-12 (int)
- `gender`: male, female, nonbinary (factor)
- `race_ethnicity`: Standard categories (factor)
- `suspensions`: Annual suspension count (int)
- `gpa`: 0-4 scale (num)
- `school_safety`: 0-10 student perception (int)
- `teacher_support`: 0-10 (int)
- `parent_involvement`: 0-10 (int)
- `school_poverty_rate`: % eligible for free lunch (num)

### 10.1 Data Loading and Quality Check

```{r}
# Load the dataset
school <- read_csv("data/school_climate.csv")

# Check the structure
glimpse(school)
```

This dataset contains student-level data nested within schools. 

### 10.2 Analyzing Discipline Disparities

Ask Claude to help you look for missing data.

```{r}

```

### 10.3 School Climate and Academic Performance

Work with Claude to find mean safety perceptions and school poverty rate.

```{r}

```

## 11. Protest Participation Patterns

**Dataset: data/protest_panel.csv**

**Description**: Three-wave panel survey on protest activity & engagement.

**Variables**:
- `respondent_id`: Panel respondent ID (int)
- `wave`: 1, 2, 3 (int)
- `age`: Baseline age (int)
- `education`: Highest degree (factor)
- `political_interest`: 0-10, time-varying (int)
- `ideology`: 1-7 scale (int)
- `participated_protest`: Protested in past 12 months (binary)
- `protest_issue`: Racial Justice, Climate, Abortion, Economy, NA (factor)
- `social_media_use`: 0-5 daily usage index (int)
- `voted_last_election`: Turnout indicator (binary)

### 11.1 Understanding Panel Data

```{r}
# Load the dataset
protest <- read_csv("data/protest_panel.csv")

# Examine panel structure
glimpse(protest)
```

Panel data follows the same individuals over time. 

### 11.2 Who Protests?

Ask Claude to help you create a profile of protesters vs non-protesters. Which group has more males? Which group is older on average?

```{r}

```

\newpage

# Segment B — Summary Statistics
```{r echo=F}
library(knitr)
library(formatR)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Non-AI Exercises

## 1. Understanding Measures of Central Tendency

### 1.1 Multiple Choice: Mean vs Median

When should you use the median instead of the mean?

a) When you want the most precise calculation
b) When the data has extreme outliers or is skewed
c) When all values are the same
d) When you have categorical data

Answer: _____

### 1.2 Fill in the Blanks: Central Tendency

Complete these statements about measures of center:

1. The _____ is the sum of all values divided by the count
2. The _____ is the middle value when data is ordered
3. The _____ is the most frequently occurring value
4. When data is skewed right, the mean is typically _____ than the median
5. The _____ is most affected by outliers

Word bank: mean, median, mode, greater, less

### 1.3 Code Detective: Grouped Summaries

What does this code calculate?

```r
approval %>%
  group_by(party_id) %>%
  summarise(
    avg_approval = mean(congress_approval, na.rm = TRUE),
    n = n()
  )
```

This code calculates: _____

## 2. Measures of Spread

### 2.1 Match: Measures of Spread

Match each measure with its definition:

**Measures:**
a) Range
b) Variance
c) Standard deviation
d) Interquartile range (IQR)

**Definitions:**
1. Square root of the variance
2. Maximum value minus minimum value
3. Average squared deviation from the mean
4. Distance between 25th and 75th percentiles

Matches: a = _____, b = _____, c = _____, d = _____

### 2.2 Multiple Choice: Standard Deviation

If approval ratings have a mean of 45% and standard deviation of 10%, approximately what percentage of responses fall between 35% and 55%?

a) 50%
b) 68%
c) 95%
d) 99%

Answer: _____ (Hint: Think about the empirical rule)

### 2.3 True or False: Spread

Mark each statement as True (T) or False (F):

_____ Variance is always positive or zero
_____ Standard deviation has the same units as the original data
_____ A larger standard deviation means data is more spread out
_____ The range is affected by outliers
_____ IQR is more robust to outliers than standard deviation

## 3. The `summarise()` Function

### 3.1 Fill in the Code: summarise()

Complete this code to calculate mean, median, and standard deviation:

```r
approval %>%
  summarise(
    mean_approval = _____(congress_approval, na.rm = TRUE),
    median_approval = _____(congress_approval, na.rm = TRUE),
    sd_approval = _____(congress_approval, na.rm = TRUE)
  )
```

Functions needed: mean, median, sd

### 3.2 Multiple Choice: NA Values

What happens when you run `mean(c(1, 2, NA, 4))`?

a) Returns 2.33 (ignores the NA)
b) Returns NA
c) Gives an error
d) Returns 2 (treats NA as 0)

Answer: _____

## 4. Real-World Applications

### 4.1 Match: Statistical Concepts

Match polling concepts with statistical terms:

**Polling Terms:**
a) Margin of error
b) Sample size
c) Confidence level
d) Poll average

**Statistical Terms:**
1. Mean of multiple measurements
2. Number of observations (n)
3. Related to standard error
4. Probability of capturing true value

Matches: a = _____, b = _____, c = _____, d = _____

### 4.2 Code Detective: Real Analysis

What question does this code answer?

```r
approval %>%
  group_by(party_id, region) %>%
  summarise(
    mean_approval = mean(congress_approval, na.rm = TRUE),
    sd_approval = sd(congress_approval, na.rm = TRUE),
    count = n()
  )
```

This analysis shows: _____

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, calculate the mean and standard deviation of congress_approval by party_id. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

## 5. Congressional Approval Analysis

**Dataset: data/congressional_approval.csv**

**Description**: Survey data on congressional approval ratings with demographics.

**Variables**:
- `respondent_id`: Unique identifier (int)
- `age`: Respondent's age (int)
- `education`: Education level (chr)
- `party_id`: Democrat, Republican, Independent (chr)
- `income_category`: Income bracket (chr)
- `region`: Geographic region (chr)
- `congress_approval`: Approval rating for Congress, 0-100 scale (dbl)

### 5.1 Initial Data Exploration

```{r}
# Load the dataset
library(tidyverse)
approval <- read_csv("data/congressional_approval.csv")

# Examine the data
glimpse(approval)
summary(approval$congress_approval)
```

Use Claude to calculate and interpret basic summary statistics for congressional approval ratings.

### 5.2 Approval by Party

Work with Claude to calculate mean and median approval ratings by party affiliation. What patterns do you observe? Are there differences in how spread out opinions are within each party?

### 5.3 Regional Variations

Ask Claude to help you explore how congressional approval varies by region. Which regions show the highest and lowest approval? Which have the most consensus (lowest standard deviation)?

### 5.4 Creating a Summary Report

Work with Claude to create a comprehensive summary table that shows approval statistics by both party and region. What insights emerge from this analysis?

## 6. Income and Wealth Distribution

**Dataset: data/household_panel.csv**

**Description**: Panel data tracking household economic conditions over time.

**Variables**:
- `hh_id`: Unique household identifier (int)
- `year`: Year of observation (int)
- `head_age`: Age of household head (int)
- `education_head`: Education of household head (chr)
- `num_children`: Number of children (int)
- `income`: Annual household income (dbl)
- `has_employer_insurance`: Has employer insurance (lgl)
- `self_reported_health`: Health status (chr)
- `unexpected_expense`: Can handle $400 expense (lgl)

### 6.1 Loading and Understanding the Data

```{r}
# Load the dataset
household <- read_csv("data/household_panel.csv")

# Check the structure
glimpse(household)
```

This is panel data (same households observed over multiple years). Ask Claude to help you understand the income and wealth distributions.

### 6.2 Income and Health

Work with Claude to explore the relationship between household income and self-reported health status. Do higher-income households report better health? Calculate appropriate summary statistics.

### 6.3 Financial Security

Ask Claude to help you analyze financial security using the unexpected_expense variable. What percentage of households can handle a $400 emergency expense? How does this vary by income and education?

## 7. Voter Turnout Patterns

**Dataset: data/voter_turnout_simple.csv**

**Description**: State-level voter turnout data from recent elections.

**Variables**:
- `state`: State name (chr)
- `turnout_2020`: Turnout percentage in 2020 (dbl)
- `turnout_2016`: Turnout percentage in 2016 (dbl)
- `population_millions`: State population in millions (dbl)

### 7.1 Turnout Overview

```{r}
# Load the dataset
turnout <- read_csv("data/voter_turnout_simple.csv")

# Explore the data
glimpse(turnout)
```

Work with Claude to calculate summary statistics for voter turnout in both 2016 and 2020. What changed between elections?

### 7.2 State-Level Patterns

Ask Claude to help you identify which states had the highest and lowest turnout in both elections. How consistent are state rankings between 2016 and 2020?

### 7.3 Turnout Changes

Work with Claude to calculate how turnout changed from 2016 to 2020 for each state. Which states saw the biggest increases or decreases?

## 8. Legislative Productivity

**Dataset: data/rollcalls.csv**

**Description**: Individual member votes on congressional bills.

**Variables**:
- `congress`: Congress number (int)
- `bill_id`: Unique bill identifier (int)
- `member_id`: Member identifier (int)
- `party`: Political party (chr)
- `ideology`: Member ideology score (dbl)
- `bill_ideology`: Bill ideology score (dbl)
- `vote`: Vote choice (chr)
- `district_partisanship`: District lean (dbl)

### 8.1 Understanding Voting Patterns

```{r}
# Load the dataset
rollcalls <- read_csv("data/rollcalls.csv")

# Look at the data
glimpse(rollcalls)
```

Use Claude to calculate basic statistics about congressional voting patterns, including passage rates and participation.

### 8.2 Party Voting Patterns

Work with Claude to analyze how members vote based on their party affiliation. Do Democrats and Republicans vote differently on bills?

### 8.3 Ideology and Voting

Ask Claude to help you explore how member ideology relates to voting patterns. Do more extreme members vote differently than moderates?

\newpage
